---
title: "Comparisson of DE genes in DSS colitis vs Human colitis"
subtitle: "Czarnewski et al 2019"
author: "Paulo Czarnewski"
date: "30 May 2019"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    toc_depth: 4
    number_sections: true
    theme: flatly
    highlight: tango
    self_contained: false
    keep_md: false
editor_options: 
  chunk_output_type: inline
---



# Load Packages

```{r,echo=TRUE}
#Installing required packages
source("./support_functions.R")
inst_packages(c("readxl","data.table","biomaRt","RColorBrewer","rafalib","enrichR","venn"))


#Defining color paletes
palete <- RColorBrewer::brewer.pal(8,"Set2")
mypalette2 <- c(brewer.pal(n = 8,name = "Set2"),brewer.pal(n = 9,name = "Set1"))
```


***
# Loading Human dataset

```{r results="hide", message=F,warning=FALSE}
UC_DEgenes <- data.frame(readxl::read_xlsx("/Users/Czarnewski/Box Sync/Paulo_data/20170913 - Human UC treatment prediction/Treatment naive UC/jjx139_suppl_supplementary_table_1.xlsx",skip = 1))
rownames(UC_DEgenes) <- UC_DEgenes[,1]
UC_DEgenes <- as.matrix(UC_DEgenes[,2:3])
```



# Loading Mouse-human homologs

Since we used the gene annotation from JAX in part of the anlaysis, it is likely that the list of differentially expressed genes might slightly vary depending on the updates in their database. For this reason, we also provided the list of genes obtained during our initial anlaysis in order to produce the exact same figures as in the paper.

```{r, results="hide", message=F, fig.width=10,warning=F,fig.height=3.3,fig.show="hold",warning=FALSE}
#A list of homolog genes can be found in:
#homologs <- fread('http://www.informatics.jax.org/downloads/reports/HMD_HumanPhenotype.rpt')
#homologs <- data.frame("mouse"=homologs$V5,"human"=homologs$V1)

#Load the homolog gene list version downloaded in 2018-06-15 used in the paper.
homologs <- read.csv("../processed_files/homologs.csv",row.names = 1)
head(homologs)
```


# Loading Mouse dataset

```{r, results="hide", message=F, fig.width=10,warning=F,fig.height=3.3,fig.show="hold",warning=FALSE}
#Loading mouse data
mouse_DEgenes <- read.csv("../processed_files/DEgenes.csv",header = T,row.names = 1)
mouse_modules <- read.csv("../processed_files/gene_modules.csv",header = T,row.names = 1)
```


# Plotting venn diagram for genes in common

```{r, results="hide", message=F, fig.width=10,warning=F,fig.height=4,fig.show="hold",warning=FALSE}
mouse_homologs <- rownames(mouse_DEgenes)
mouse_homologs <- mouse_homologs[!(rownames(mouse_DEgenes) %in% homologs[,1] )]
mouse_homologs <- c(mouse_homologs, as.character(homologs[homologs[,1] %in% rownames(mouse_DEgenes),2]))

mypar(mar=c(2,2,4,2))
input  <- list(human_DEG=as.character(rownames(UC_DEgenes)),
               mouse_DEG=mouse_homologs)
a <- attr(venn(input,cexil = 2,cexsn = 2), "intersections")
title(main = paste0("Venn diagram for all DEGs"))
```



# Plotting venn diagram per cluster

```{r, results="hide", message=F, fig.width=10,warning=F,fig.height=10,fig.show="hold",warning=FALSE}
intersect_per_cluster <- list()

mypar(3,3,mar=c(2,2,4,2))
for(i in 1:max(mouse_modules[,1])){
  mouse_homologs_module <- rownames(mouse_DEgenes)[mouse_modules[,1] == i]
  mouse_homologs_module <- mouse_homologs_module[!(rownames(mouse_DEgenes)[mouse_modules[,1] == i] %in% homologs[,1] )]
  mouse_homologs_module <- c(mouse_homologs_module, as.character(homologs[homologs[,1] %in% rownames(mouse_DEgenes)[mouse_modules[,1] == i],2]))
  
  input  <- list(human_DEG=as.character(rownames(UC_DEgenes)),
                 mouse_DEG=mouse_homologs_module)
  a<-attr(venn(input,cexil = 1.5,cexsn = 1.5), "intersections")
  title(main = paste0("Module ",i))
  
  intersect_per_cluster[[i]] <- a$`human_DEG:mouse_DEG`
}
```

# Plotting percentages per cluster


```{r,  results="hide", message=F, fig.width=10,warning=F,fig.height=3,fig.show="hold",warning=FALSE}
temp <- sapply(intersect_per_cluster,length)
names(temp) <- paste0("module",1:max(mouse_modules[,1]))

mypar(1,2,mar=c(4,5,2,3))
barplot(temp[order(temp)],horiz=T,yaxs="i",col=mypalette2[1:max(mouse_modules[,1])][order(temp)],las=1,border=NA,xaxs="i",xlim=c(0,200),xlab='number of genes')
abline(v=0,lwd=2)
text(temp[order(temp)],(0:(length(temp)-1)*1.2)+0.7,paste0(100*round(temp[order(temp)]/sum(temp),3),"%"),pos=4)
pie(temp[order(temp,decreasing = T)],col = mypalette2[1:max(mouse_modules[,1])][order(temp,decreasing = T)],clockwise = T,border = "white",main = "Percentage of conserved gene per cluster")
```


# Performing Gene Set Analysis on the shared list between mouse and humans


```{r results="hide", message=F, warning=FALSE, fig.width=10, warning=F, fig.height=3}
pvalue_cutoff <- 0.05
no_genes_cutoff <- 3


#GO enrichment
a <- enrichr(genes=unique(unlist(intersect_per_cluster)), databases = "GO_Biological_Process_2017")[[1]]
  #remove terms with p-values > pvalue_cutoff and number of genes < no_genes_cutoff
a <- a[order(a$P.value,decreasing = F),]
a <- a[a$P.value < pvalue_cutoff & as.numeric(sapply(strsplit(a[,"Overlap"],"/"),"[[", 1)) >= no_genes_cutoff,]
  #remove terms with "positive/negative regulation"
a <- a[grep("regulation",a$Term,invert = T),]
  #if (nrow(a) > 20) { a <- a[order(a$P.value,decreasing = F),][1:20,] }
a$Term <- sapply(strsplit(a[,"Term"]," [(]GO:"),"[[", 1)


#KEGG enrichment
b <- enrichr(genes=unique(unlist(intersect_per_cluster)), databases = "KEGG_2016")[[1]]
b <- b[order(b$P.value,decreasing = F),]
b <- b[grep("_hsa05",b$Term,invert = T),]   #Remove disease-related KEGG pathways ("hsa05"")
b <- b[b$P.value < pvalue_cutoff & as.numeric(sapply(strsplit(b[,"Overlap"],"/"),"[[", 1)) >= no_genes_cutoff,]
  #if (nrow(b) > 20) { b <- b[order(b$P.value,decreasing = F),][1:20,] }
b$Term <- sapply(strsplit(b[,"Term"],"_"),"[[", 1)


#Plot enrichments
n <- 10
mypar(1,2,mar=c(2,15,2,3))
barplot(-log(a[n:1,"P.value"],10),names.arg=a[n:1,1],horiz=T,las=1,col="grey",main='GO',xlab=" ",xlim=c(0,max(-log(a[n:1,"P.value"],10))*1.2),tck=-.05,border=NA,cex.axis=1,cex.names=1)
abline(v=0,lwd=2)
barplot(-log(b[n:1,"P.value"],10),names.arg=b[n:1,1],horiz=T,las=1,col="grey",main='KEGG',xlab=" ",xlim=c(0,max(-log(b[n:1,"P.value"],10))*1.2),tck=-.05,border=NA,cex.axis=1,cex.names=1)
abline(v=0,lwd=2)
```

